The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-06-05 script/en/chapters/cryptomethods.tex
fixed warnings
CT Script: Added chap 1.1 (sec def), and added OTP in 2.2.4 (EN only yet)

\hypertarget{Kapitel_1}{}
\chapter{Encryption Procedures}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Joerg-Cornelius Schneider, May 1999; Updates Dec. 2001, Feb. 2003,
June 2005, July 2007, January 2010)

@@

The purpose of encryption \index{Encryption} is to change data in such a way
that only an authorized recipient is able to reconstruct the plaintext. This
allows us to transmit data without worrying about it getting into unauthorized
hands. Authorized recipients possess a piece of secret information -- called the
key -- which allows them to decrypt the data while it remains hidden from
everyone else.\par \vskip + 3pt

One encryption procedure has been mathematically proved to be secure, the {\em
One Time Pad}\index{One Time Pad}. However, this procedure has several
practical disadvantages (the key used must be randomly selected and must be at
least as long as the message being protected), which means that it is hardly
used except in closed environments such as for the hot wire between Moscow and
Washington.\par \vskip + 3pt

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do so, and these procedures can therefore be
considered (practically) secure.\par \vskip + 3pt

The book of Bruce Schneier \cite{cm:Schneier1996cm} offers a very good overview of
the different algorithms. We basically distinguish between symmetric and
asymmetric encryption procedures.

@@

% --------------------------------------------------------------------------
\subsection{New results about cryptanalysis of AES}
\label{NeueAES-Analyse}
\index{Cryptanalysis}
\index{AES}

@@

>>>>

% HACK to fix warning "destination with the same identifier .. has already been used, ...":
\makeatletter \renewcommand{\thepage}{~\csname @arabic\endcsname \c@page~} \makeatother
\hypertarget{Kapitel_1}{}
\chapter{Encryption Procedures}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Joerg-Cornelius Schneider, May 1999; Updates Dec. 2001, Feb. 2003,
June 2005, July 2007, January 2010, June 2012)

@@

The purpose of encryption \index{Encryption} is to change data in such a way
that only an authorized recipient is able to reconstruct the plaintext. This
allows us to transmit data without worrying about it getting into unauthorized
hands. Authorized recipients possess a piece of secret information -- called the
key -- which allows them to decrypt the data while it remains hidden from
everyone else.%\par \vskip + 3pt

First we present the ideas how the security of cryptosystems is defined.

% --------------------------------------------------------------------------
\hypertarget{cm_Section_Security_Definitions}{}
\section[Security definitions]{Security definitions}
\label{cm_Section_Security_Definitions}

Modern cryptography is heavily based on mathematical theory and computer
science practice. Cryptographic algorithms are designed around computational
hardness assumptions, making such algorithms hard to break in practice by any
adversary.

\vskip +3pt
Depending on the adversary's capabilities there are mainly two basic notations
of security distinguished in literature
(see e.g. {\em Contemporary Cryptography} \cite{cm:Oppliger2011}):

\begin{itemize}

\item {\bf Computational, conditional or practical security}\\
A cipher is computationally secure if it is theoretically possible to break such
a system but it is infeasible to do so by any known practical means. Theoretical
advances (e.g., improvements in integer factorization algorithms) and faster
computing technology require these solutions to be continually adapted.

Even using the best known algorithm for breaking it will require so much
resources (e.g., 1,000,000 years) that practically the cryptosystem is secure.

So this concept is based on assumptions of the adversary's limited computing
power and the current state of science.

\item {\bf Information-theoretical or unconditional security}\\
A cipher is unconditionally secure if it is secure no matter how much resources
(time, space) the attacker has. So even in the case where an adversary has
unlimited resources for solving a problem e.g. breaking a cipher. An adversary
has not sufficient information for doing that because he is unable to gain any
meaningful data from a ciphertext.

There exist information-theoretically secure schemes that provably cannot be
broken even with unlimited computing power -- an example is the
{\em One Time Pad}\index{One Time Pad} (OTP\index{OTP}).

As the OTP is information-theoretically secure it derives its security
solely from information theory and is secure even with unlimited computing
power at the adversary's disposal. However, OTP has several
practical disadvantages (the key used must be randomly selected and must be at
least as long as the message being protected), which means that it is hardly
used except in closed environments such as for the hot wire between Moscow and
Washington.%\par \vskip + 3pt

\vskip +3pt
Two more concepts are sometimes used:

\begin{itemize}

\item {\bf Provable security}
This means that breaking such a cryptographic system is as difficult as
solving some supposedly difficult problem e.g. discrete logarithm computation,
discrete square root computation, very large integer factorization.

Example: Currently we know that RSA is at most as difficult as factorization,
but we cannot prove that its exactly as difficult as factorization. So RSA
has no proven minimum security. Or in other words: We cannot prove, that if
RSA (the cryptosystem) is broken, that then factorization (the hard
mathematical problem) can be solved.

The Rabin cryptosystem was the first cryptosystem which could be proven to be
computationally equivalent to a hard problem.

\item {\bf Ad-hoc security}
A cryptographic system has this security feature if it is not worth to try to
break into such a system because of inadequate price of data with comparison to
price of work needed to do so. Or an attack can't be done in sufficiently short
time (see {\em Handbook of Applied Cryptography} \cite{cm:Menezes2001}).

Example: This may apply if a message relevant for the stock market will be
published tomorrow and you need a year to break it.

\vskip +3pt
For good procedures used today the time taken to break them is so long that
it is practically impossible to do so, and these procedures can therefore be
considered (practically) secure.%\par \vskip + 3pt

The book of Bruce Schneier \cite{cm:Schneier1996cm} offers a good overview of
the different algorithms. We basically distinguish between symmetric and
asymmetric encryption procedures.

@@

% HACK to fix warning "destination with the same identifier .. has already been used, ...":
\makeatletter \renewcommand{\thepage}{\csname @arabic\endcsname \c@page} \makeatother
% --------------------------------------------------------------------------
\subsection{New results about cryptanalysis of AES}
\label{NeueAES-Analyse}
\index{Cryptanalysis}
\index{AES}

@@

\bibitem[Menezes2001]{cm:Menezes2001} \index{Menezes 2001}
    Alfred J. Menezes, Paul C. van Oorschot, Scott A. Vanstone \\
    {\em Handbook of Applied Cryptography}, 
    CRC Press 1997, 5th printing 2001,\\
    \url{http://www.cacr.math.uwaterloo.ca/hac/} (Errata last updated July 24, 2011).

### 2012-06-05 script/en/chapters/paper_and_pencil.tex
CT Script: Added chap 1.1 (sec def), and added OTP in 2.2.4 (EN only yet)

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes to this chapter describe how the methods can be
performed using CrypTool CrypTool 1.
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code using the computer algebra system Sage\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\begin{itemize}
\item {\bf Vigen\`ere}\footnote{%
   In CrypTool\index{CrypTool} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Vigen\`ere}.}  
   \cite{pp:Singh1999}\index{Vigen\`ere}: 
   Each plaintext character is encrypted
   with a different ciphertext alphabet that is determined by the characters of
   a keyword (the so-called Vigen\`ere-Tableau serves auxiliary means).
   If the plaintext is longer than the key, the latter is repeated.

@@

\item {\bf Superposition}\index{Superposition}
   \begin{itemize}
       \item {\bf Book cipher}: A keytext (for example out of a book) is added
          to the plaintext.
       \item {\bf Superposition with numbers}: A sequence or a number of 
          sufficient length (for example pi) is added.
   \end{itemize}

>>>>

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes to this chapter describe how the methods can be
performed using CrypTool 1 (CT1), CrypTool 2 (CT2) or JCrypTool (JCT).
\index{CT1} \index{CT2} \index{JCT}
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code using the computer algebra system Sage\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\begin{itemize}
\item {\bf Vigen\`ere}\footnote{%
   In CT1\index{CT1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Vigen\`ere}.}  
   \cite{pp:Singh1999}\index{Vigen\`ere}: 
   Each plaintext character is encrypted
   with a different ciphertext alphabet that is determined by the characters of
   a keyword (the so-called Vigen\`ere tableau serves auxiliary means).
   If the plaintext is longer than the key, the latter is repeated.

@@

\item {\bf Superposition}\index{Superposition}
   \begin{itemize}
       \item {\bf Running-key cipher}: A keytext (for example out of a book) is added
          to the plaintext.

       \item {\bf Superposition with numbers}: A sequence or a number of 
          sufficient length (for example pi) is added.

       \item {\bf One Time Pad (OTP)}\footnote{%
          CT1\index{CT1} offers this method under the menu
          {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
          \textbackslash{} Vernam / OTP}.}\index{One Time Pad}\index{OTP}:
          A sequence
          of bytes is XORed byte-by-byte to the plaintext. This is a
          generalization of Vigen\`ere's mechanism and it was the first
          information-theoretically secure scheme (see
          chapter~\ref{cm_Section_Security_Definitions}
          ``\nameref{cm_Section_Security_Definitions}'').
          To fulfill this claim the pad must be random and it must be used
          only once (to eliminate any semblance of pattern from ciphertext).\\
          Reason: Given ciphertext C, plaintext P, pad K and two plaintexts
          encrypted with the same key: C1 = P1 + K;  C2 = P2 + K; \\
          Thus, C1 + C2 = (P1 + K) + (P2 + K) = P1 + P2; \\
          which effectively could leak the plaintexts\footnote{%
          In JCT\index{JavaCrypTool}\index{JCT} you can play with an automatic
          cryptanalysis of running-key ciphertexts using the Viterbi analysis
          under the menu {\bf Visuals \textbackslash{} Viterbi}.}.
   \end{itemize}
